# tAge Replication

## #1 Python and Directories

```{r}
dir_python=system("where python", intern=TRUE)

Sys.setenv(
  RETICULATE_PYTHON = utils::shortPathName(dir_python[1])
)
```

## #2 Load Packages

```{r}
library(tAge)
library(Biobase)
library(ggplot2)
library(here)
library(readxl)
```

## #3 Import Data

### Expression

```{r}
expression=read.csv(here("data","raw","Expression_data_absolute_rodents_Scaled.csv"),
  row.names = 1,
  check.names = FALSE
)
```

### Metadata

#### *Tabela só com ITPs*

```{r}
metadata <- read.csv(here("results","table_Zenodo.csv"),
    row.names = 1,
    check.names = FALSE
)
```

#### *Tabela completa*

```{r}
metadata <- as.data.frame(read_excel(here("data","raw","Data_annotation_absolute_rodents.xlsx"),
  na = c("", "NA"), guess_max = 10000
))

rownames(metadata) <- as.character(metadata[[1]])
```

### Create New Column for Cohort_Lab

```{r}
metadata$Cohort_Lab <- paste(metadata$Cohort, metadata$Lab, sep = "_")
```

## #4 Common Samples

### Select Common Samples

```{r}
common = intersect(colnames(expression),rownames(metadata))
paste0(length(common)," samples in common")
```

### Discard Different Samples

```{r}
expression <- expression[, common]

metadata <- metadata[common, ]
```

## #5 Create Expression Set

```{r}
eset = make_ExpressionSet(
  expression,
  metadata
)
```

## #6 Gene alignment and subtracting control values

### Gene Alignment

```{r}
gene_list <- tAge:::load_gene_list()

e <- tAge:::.align_to_gene_list(eset, gene_list)
```

### Division by Groups

There will be 4 different groups to calculate the mean control value:

- All 55 controls

- By Cohort and Sex

- By Lab

- By Cohort and Lab

### Subtract_by function

```{r}
control_column <- "Intervention"
control_label  <- "Control"

subtract_by <- function(eset, group_cols = NULL, control_col, control_lab) {
  mat  <- Biobase::exprs(eset)
  meta <- Biobase::pData(eset)
  
  is_control <- meta[[control_col]] %in% control_lab
  group_id <- if (is.null(group_cols)) {
    rep("all", ncol(mat))
  } else {
    do.call(paste, c(meta[group_cols], sep = "_"))
  }
  
  out <- mat
  has_controls <- rep(TRUE, ncol(mat))
  
  for (g in unique(group_id)) {
    in_group   <- group_id == g
    ctrl_group <- in_group & is_control
    if (sum(ctrl_group) == 0) {
      has_controls[in_group] <- FALSE
      next
    }
    control_median <- apply(mat[, ctrl_group, drop = FALSE], 1, median, na.rm = TRUE)
    out[, in_group] <- mat[, in_group, drop = FALSE] - control_median
  }
  
  if (any(!has_controls)) {
    message(sum(!has_controls), " samples dropped (no controls in their group)")
  }
  
  Biobase::exprs(eset) <- out
  eset[, has_controls]
}
```

### Control Subtraction by groups

```{r}
groups <- list(
  All_Controls = NULL,
  Cohort_Sex   = c("Cohort", "Sex"), # Esta linha acho que está "mal", poderia ser "Group"
  Lab          = "Lab",
  Cohort_Lab   = "Cohort_Lab"
)

eset_by_group <- lapply(groups, function(cols) {
  subtract_by(e, cols, control_column, control_label)
})
```

### Subtract Controls from Expected_Hazard.log10

```{r}
center_by <- function(values, meta, group_cols = NULL, control_col, control_lab) {
  is_control <- meta[[control_col]] %in% control_lab
  group_id <- if (is.null(group_cols)) {
    rep("all", length(values))
  } else {
    do.call(paste, c(meta[group_cols], sep = "_"))
  }
  
  out <- values
  for (g in unique(group_id)) {
    in_group <- group_id == g
    control_median <- median(values[in_group & is_control], na.rm = TRUE)
    out[in_group] <- values[in_group] - control_median
  }
  out
}
```

## #7 Find the Models

```{r}
model_dir <- here("data","raw")

model_files <- list.files(model_dir, pattern = "\\.pkl$", full.names = TRUE)
basename(model_files)
```

## #8 Run the Models

```{r}
references <- c(
  Mortality_Rodents_Liver = "Expected_Hazard.log10",
  Mortality_Mouse_Multitissue = "Expected_Hazard.log10",
  Mortality_Rodents_Multitissue = "Expected_Hazard.log10"
)

results_list <- list()

for (group_name in names(eset_by_group)) {
  e_sub <- eset_by_group[[group_name]]
  meta_sub <- Biobase::pData(e_sub)
  
  for (model_name in names(references)) {
    model_path <- model_files[grepl(model_name, basename(model_files), fixed = TRUE)]
    if (length(model_path) != 1) {
      warning(model_name, ": ", length(model_path), " matching files, skipped")
      next
    }
    
    pred <- predict_tAge(
      list(scaled_diff = e_sub),
      list(scaled_diff = model_path),
      species = "mouse",
      mode ="EN"
    )
    
    pred_cols <- setdiff(colnames(pred), colnames(meta_sub))
    pred_col  <- pred_cols[!grepl("_sd$", pred_cols)][1]
    
    stopifnot(all(colnames(e_sub) %in% rownames(pred)))
    obtained <- pred[colnames(e_sub), pred_col]
    expected <- center_by(
      as.numeric(meta_sub[[references[[model_name]]]]),
      meta_sub,
      groups[[group_name]],
      control_column,
      control_label
    )
    difference <- obtained - expected
    
    block <- data.frame(
      Obtained   = obtained,
      Expected_Centered = expected,
      Difference = difference,
      Mean_Abs_Difference = NA_real_,
      Max_Abs_Difference  = NA_real_,
      SD_Difference       = NA_real_,
      Pearson  = NA_real_,
      Spearman = NA_real_,
      row.names = colnames(e_sub)
    )
    
    # summary metrics: only the first row of each block
    block$Mean_Abs_Difference[1] <- mean(abs(difference), na.rm = TRUE)
    block$Max_Abs_Difference[1]  <- max(abs(difference), na.rm = TRUE)
    block$SD_Difference[1]       <- sd(difference, na.rm = TRUE)
    block$Pearson[1]  <- cor(obtained, expected, use = "complete.obs", method = "pearson")
    block$Spearman[1] <- cor(obtained, expected, use = "complete.obs", method = "spearman")
    
    names(block) <- paste(group_name, model_name, names(block), sep = "__")
    results_list[[paste(group_name, model_name, sep = "__")]] <- block
  }
}
```

## #9 Results Table

```{r}
aligned_blocks <- lapply(results_list, function(b) {
  out <- b[rownames(metadata), , drop = FALSE]
  rownames(out) <- NULL
  out
})

expected_centered <- sapply(names(groups), function(g) {
  center_by(
    as.numeric(metadata[["Expected_Hazard.log10"]]),
    metadata,
    groups[[g]],
    control_column,
    control_label
  )
})
colnames(expected_centered) <- paste0(names(groups), "__Expected_Hazard.log10_centered")

results_table <- data.frame(
  Sample = rownames(metadata),
  metadata,
  expected_centered,
  " " = NA,
  do.call(cbind, unname(aligned_blocks)),
  check.names = FALSE,
  row.names = NULL
)

write.csv(results_table,
          here("experiments","replication","tAge_replication_results.csv"),
          na="",
          row.names= FALSE
          )
```

## #10 Graphs

```{r}
plot_data <- do.call(rbind, lapply(names(aligned_blocks), function(key) {
  parts <- strsplit(key, "__")[[1]]
  data.frame(
    Control_Group = parts[1],
    Model    = parts[2],
    Sample   = rownames(metadata),
    Obtained = aligned_blocks[[key]][[paste(key, "Obtained", sep = "__")]],
    Expected = aligned_blocks[[key]][[paste(key, "Expected_Centered", sep = "__")]],
    Lab      = metadata$Lab
  )
}))

# Pearson r per panel, shown in the corner
panel_stats <- do.call(rbind, lapply(
  split(plot_data, list(plot_data$Control_Group, plot_data$Model), drop = TRUE),
  function(d) {
    data.frame(
      Control_Group = d$Control_Group[1],
      Model = d$Model[1],
      label = sprintf("r = %.2f", cor(d$Obtained, d$Expected, use = "complete.obs"))
    )
  }
))

scatter_plot <- ggplot(plot_data, aes(x = Expected, y = Obtained, colour = Lab)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey50") +
  geom_point(alpha = 0.7, size = 1.8, na.rm = TRUE) +
  geom_text(
    data = panel_stats, aes(label = label),
    x = -Inf, y = Inf, hjust = -0.1, vjust = 1.5,
    inherit.aes = FALSE, size = 3.5
  ) +
  facet_grid(Control_Group ~ Model) +
  labs(
    x = "Expected hazard (log10)",
    y = "Obtained hazard (log10)",
    title = "Obtained vs expected hazard by control group and model"
  ) +
  theme_bw()

scatter_plot

ggsave(here("experiments","replication","tAge_scatterplots.png"),
  scatter_plot, width = 14, height = 11, dpi = 300
)
```
